# Results


## How do suicide death counts vary across different age groups, genders, regions, and ethnicities in New York State?
```{r, message=FALSE}
library(reshape2)
library(dplyr)
library(ggplot2)
library(RColorBrewer)

death_data <- read.csv("Death_data_NY.csv")
death_data <- death_data %>% 
  filter(Age.Group != "Unknown")

aggregated_data <- death_data %>%
   filter(Age.Group != 'Total') %>%
  group_by(Race.or.Ethnicity, Age.Group) %>%
  summarise(Suicide.Deaths = sum(Suicide.Deaths, na.rm = TRUE))

# Use melt from reshape2 to reshape the data for heatmap
data_melted <- melt(aggregated_data, id.vars = c('Race.or.Ethnicity', 'Age.Group'))

# Generate the heatmap
ggplot(data_melted, aes(x = Age.Group, y = Race.or.Ethnicity, fill = value)) +
    geom_tile() +
    scale_fill_gradientn(colors = brewer.pal(9, "Blues")) +
    labs(title = "Heatmap of Suicide Counts by Age Group and Ethnicity (18 years)",
         x = "Age Group",
         y = "Race or Ethnicity",
         fill = "Suicide Counts") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
**Comment:** The heatmap provides a visual representation of suicide counts by age group and ethnicity, with darker colors indicating higher numbers. It shows that White Non Hispanic individuals indeed have a notably higher incidence of suicide across most age groups, particularly in the middle-aged brackets. Other ethnic groups, such as Hispanic and Black Non Hispanic, have lower counts, which are more evenly distributed across age groups. The "Not Stated" category suggests a lack of data or unreported ethnicity in some cases. The concentration of darker shades in specific age ranges may highlight critical periods where interventions could be most impactful.  


```{r, message=FALSE}
most_recent_year <- max(death_data$Year)
data_recent_year <- subset(death_data, Year == most_recent_year & Age.Group != 'Total')

# Summarize suicide deaths by age group, sex, and race or ethnicity
suicide_summary <- data_recent_year %>%
  group_by(Age.Group, Sex, Race.or.Ethnicity) %>%
  summarise(Suicide_Deaths = sum(Suicide.Deaths, na.rm = TRUE))

# Create a facet grid plot
ggplot(suicide_summary, aes(x = Age.Group, y = Suicide_Deaths, fill = Sex)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~ Race.or.Ethnicity, scales = "free_y") +
  labs(title = "Suicide Deaths by Age, Gender, and Ethnicity (2020)",
       x = "Age Group",
       y = "Number of Suicide Deaths",
       fill = "Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
**Comment:** The bar charts depict trends of suicide deaths across different age groups and ethnicities for the year 2020, with gender comparisons. Notably, the "White Non Hispanic" category has the highest numbers, especially among middle-aged individuals. Men in this group appear particularly affected, as seen by the higher bars in comparison to women. Other ethnic groups show fewer suicide deaths, but still present a peak in middle age. The data suggests that middle age is a critical period for suicide risk, especially among White Non Hispanic men.  

```{r, message=FALSE}
aggregated_data_5 <- death_data %>%
  filter(Age.Group != 'Total') %>%
  group_by(Year, Sex) %>%
  summarise(
    Suicide.Deaths = sum(Suicide.Deaths, na.rm = TRUE),
  ) %>%
  ungroup() # Remove the grouping

ggplot(aggregated_data_5, aes(x = Sex, y = Suicide.Deaths, fill = Sex)) +
  geom_boxplot(outlier.shape = NA) +  # Hide outliers
  geom_jitter(width = 0.2, alpha = 0.5, size = 2) +  # Add data points with some transparency
  scale_fill_brewer(palette = "Pastel1") +  # A soft color palette
  labs(
    title = "Annual Suicide Death Counts by Gender (2003-2020)",
    subtitle = "New York State data",
    x = "Gender",
    y = "Suicide Death Counts",
    fill = "Gender"
  ) +
  theme_minimal() +  # A clean theme
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "bottom",
    legend.title.align = 0.5,
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold")
  ) +
  scale_x_discrete(labels = c("F" = "Female", "M" = "Male"))
```
**Comment:** The box plot illustrates the annual suicide death counts by gender in New York State from 2003 to 2020. It reveals that males have a significantly higher number of suicide deaths annually compared to females, as indicated by the higher median value and larger spread of the data for males. The data points represented by dots show variation from year to year, with several outliers indicating years with particularly high or low counts. The consistent position of the box for males at a higher count than females suggests that this is a long-standing trend over the examined period. This visual summary underscores a gender disparity in suicide rates, with males being more affected than females.

```{r, message=FALSE}
aggregated_data_6 <- death_data %>%
  filter(Age.Group != 'Total') %>%
  group_by(Year, Sex) %>%
  summarise(
    Suicide.Deaths = sum(Suicide.Deaths, na.rm = TRUE),
    Firearm.Deaths = sum(Firearm.Deaths, na.rm = TRUE),
    Alcohol.Related.Deaths = sum(Alcohol.Related.Deaths, na.rm = TRUE)
  ) %>%
  ungroup() # Remove the grouping


ggplot(aggregated_data_6, aes(x = Suicide.Deaths, y = Firearm.Deaths, color = Sex)) +
  geom_point(size = 3, alpha = 1) +  # Points with slight transparency
  theme_minimal() +
  labs(
    title = "Scatter Plot of Suicide vs. Firearm Deaths by Gender",
    x = "Suicide Deaths",
    y = "Firearm Deaths",
    color = "Gender"
  ) +
  scale_color_manual(values = c("F" = "pink", "M" = "lightblue"))

```


```{r, message=FALSE}
library(rgl)

mycolors <- c('pink', 'lightblue')

aggregated_data_6$Sex <- factor(aggregated_data_6$Sex, levels = c('F', 'M'))

aggregated_data_6$color <- mycolors[as.numeric(aggregated_data_6$Sex)]

plot3d( 
  x=aggregated_data_6$Suicide.Deaths, 
  y = aggregated_data_6$Firearm.Deaths, 
  z=aggregated_data_6$Alcohol.Related.Deaths, 
  col = aggregated_data_6$color, 
  type = 's', 
  radius = 50,
  title = "1",
  xlab="Suicide Deaths", ylab="Firearm Deaths", zlab="Alcohol Related Deaths")

# Add a title to the plot
title3d(main = "3D Scatter Plot of Death Counts (by gender)", line = 2)

legend3d("topright", legend = levels(aggregated_data_6$Sex), 
         col = mycolors, pch = 16, cex = 0.75)

rglwidget()
```

## Has there been a significant change in the number of suicide, firearm or alcohol-related deaths in New York State over the years?
```{r, message=FALSE}
library(ggplot2)
library(dplyr)

death_data <- read.csv("Death_data_NY.csv")
death_data <- death_data %>% 
  filter(Age.Group != "Unknown")

# Group data by year and summarize
annual_summary <- death_data %>%
  group_by(Year) %>%
  filter(Age.Group != 'Total') %>% 
  summarise(Suicide_Total = sum(Suicide.Deaths, na.rm = TRUE),
            Alcohol_Related_Total = sum(Alcohol.Related.Deaths, na.rm = TRUE),
            Firearm_Total = sum(Firearm.Deaths, na.rm = TRUE))

# Plotting Suicide, Alcohol-Related, and Firearm Deaths Over Years
ggplot() +
  geom_line(data = annual_summary, aes(x = Year, y = Suicide_Total, color = "Suicide Deaths")) +
  geom_line(data = annual_summary, aes(x = Year, y = Alcohol_Related_Total, color = "Alcohol-Related Deaths")) +
  geom_line(data = annual_summary, aes(x = Year, y = Firearm_Total, color = "Firearm Deaths")) +
  labs(title = "Suicide and Alcohol-Related Deaths in New York State Over the Years",
       x = "Year",
       y = "Total Deaths") +
  scale_color_manual(values = c("Suicide Deaths" = "blue", "Alcohol-Related Deaths" = "red", "Firearm Deaths" = "green")) +
  theme_minimal() +
  theme(legend.title = element_blank())
```


```{r, message=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
# Reshape the data to long format
long_data <- death_data %>%
  filter(Age.Group != 'Total') %>% 
  pivot_longer(cols = c("Suicide.Deaths", "Firearm.Deaths", "Alcohol.Related.Deaths"),
               names_to = "Category", values_to = "Deaths")

# Create a stacked bar chart
ggplot(long_data, aes(x = as.factor(Year), y = Deaths, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Stacked Histogram of Deaths by Category Over the Years",
       x = "Year",
       y = "Number of Deaths",
       fill = "Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels
```







## Which regions in New York State have the highest counts of suicide, alcohol-related, and firearm deaths?
```{r, message=FALSE, warning=FALSE}
library(ggmosaic)
library(tidyr)
library(dplyr)

aggregated_data_4 <- death_data %>%
  filter(Age.Group != 'Total') %>%
  group_by(Region, Sex) %>%
  summarise(
    Suicide.Deaths = sum(Suicide.Deaths, na.rm = TRUE),
    Firearm.Deaths = sum(Firearm.Deaths, na.rm = TRUE),
    Alcohol.Related.Deaths = sum(Alcohol.Related.Deaths, na.rm = TRUE)
  ) %>% ungroup() # Remove the grouping

data_long <- aggregated_data_4  %>%
  pivot_longer(cols = c("Suicide.Deaths", "Firearm.Deaths", "Alcohol.Related.Deaths"),
               names_to = "DeathType",
               values_to = "Count")
data_long$Count <- as.numeric(data_long$Count)


# Create the mosaic plot using the long format data
ggplot(data_long) +
  geom_mosaic(aes(x = product(Sex, Region), fill = DeathType, weight = Count)) +
  theme_minimal() +
  labs(title = "Mosaic Plot of Death Counts by Gender and Region (18 years)",
       x = "Region",
       y = "Gender",
       fill = "Type of Death") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x labels for clarity


```

```{r, message=FALSE}
library(RColorBrewer)

# Adding population data for NYC and ROS
nyc_population <- 8770000 # replace with the actual population of New York City
ros_population <- 19340000 - 8770000 # replace with the actual population of the Rest of State

df_2020 <- death_data %>%
  filter(Year == 2020)
# Add the population data to the dataframe
df_2020$Population <- ifelse(df_2020$Region == "NYC", nyc_population, ros_population)

# Calculate death rates per 100,000 population
df_rates_2020 <- df_2020 %>%
  pivot_longer(cols = c("Suicide.Deaths", "Firearm.Deaths", "Alcohol.Related.Deaths"), 
               names_to = "DeathType", 
               values_to = "Count") %>%
  mutate(Rate = (Count / Population) * 100000)

ggplot(df_rates_2020, aes(x = Region, y = Rate, fill = DeathType)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  theme_minimal() +
  labs(title = "Death Rates by Type and Region in New York State (2020)",
       x = "Region",
       y = "Death Rate (per 100,000 Population)",
       fill = "Type of Death") +
  scale_fill_brewer(palette = "Paired")

```


## How have mortality patterns by age and race, specifically related to suicide, firearm, and alcohol-related deaths, changed in New York State over the years?

```{r, message=FALSE}
library(ggplot2)
library(dplyr)
library(ggridges)


# Calculate the total death count
death_data$Total_Deaths <- rowSums(death_data[, c("Suicide.Deaths", "Alcohol.Related.Deaths", "Firearm.Deaths")], na.rm = TRUE)

# Reshape the data
long_data <- death_data %>%
  filter(Age.Group != 'Total') %>% 
  select(Year, Age.Group, Total_Deaths) %>%
  group_by(Year, Age.Group) %>%
  summarise(Total_Deaths = sum(Total_Deaths, na.rm = TRUE)) %>%
  ungroup()

# Create the ridgeline plot
ggplot(long_data, aes(x = Total_Deaths, y = Age.Group, fill = Age.Group)) +
  geom_density_ridges() +
  labs(title = "Ridgeline Plot of Total Death Counts Across Age Groups",
       x = "Total Death Counts",
       y = "Age Group") +
  scale_fill_brewer(palette = "Set3") + # Use a discrete color scale
  theme_ridges(grid = TRUE)
```


```{r, message=FALSE}
library(GGally)
library(dplyr)

aggregated_data_3 <- death_data %>%
  filter(Age.Group != 'Total') %>%
  group_by(Race.or.Ethnicity) %>%
  summarise(
    Suicide.Deaths = sum(Suicide.Deaths, na.rm = TRUE),
    Firearm.Deaths = sum(Firearm.Deaths, na.rm = TRUE),
    Alcohol.Related.Deaths = sum(Alcohol.Related.Deaths, na.rm = TRUE)
  ) %>%
  ungroup() 

normalized_data <- as.data.frame(scale(aggregated_data_3[,-1]))

colnames(normalized_data) <- colnames(aggregated_data_3)[2:ncol(aggregated_data_3)]

normalized_data$Race.or.Ethnicity <- aggregated_data_3$Race.or.Ethnicity

ggparcoord(normalized_data,
           columns = 1:3, 
           groupColumn = 'Race.or.Ethnicity', # Grouping by race or ethnicity
           scale = "globalminmax", # Normalize the variables globally
           mapping = aes(color = Race.or.Ethnicity)) +
  theme_minimal() +
  labs(title = "Parallel Coordinate Plot of Death Counts by Race (18 Years)",
       x = "Type of Death",
       y = "Normalized Death Counts") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_brewer(palette = "Set1")
```



